Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-27 manuscript/manuscript.tex
added Juhye refs to text

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs\jdbcomment{citation, I think Juhye has something she cites that looks at really diverged HAs}.

@@

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function\jdbcomment{citation, Juhye can probably help}.
We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty probably lies in the length of the long branches separating the subtypes rather than in the tree topology itself.

>>>>

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

@@

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty probably lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
